Box 1 - TDT Curves

### 2023-09-27 ms/ms.qmd
Updated Box 1.

Phylogenetic comparative approaches could be leveraged to help fill these knowledge gaps because closely related species are often more likely to share similar traits owing to their evolutionary history (phylogenetic niche conservatism) [@Losos2008]. If heat tolerance is linked to the phylogeny, shared characteristics that make species more or less heat tolerant will not be randomly distributed across the phylogeny but biased by species groups [@McKinney1997]. We can use evolutionary relatedness to predict species thermal sensitivities and such insights can be used to inform policy and conservation related decisions making for species with little to no data.

@@

##### Evolutionary relationships can help understand thermal tolerance in species lacking data
The model predicted known data reasonably well for most traits (correlation between known values and values estimated from the model: $z^{TDT}$ – `r cor.test(check_z$trait, check_z$est)$estimate`, 95% CI: `r cor.test(check_z$trait, check_z$est)$conf.int[1]` to `r cor.test(check_z$trait, check_z$est)$conf.int[2]`, p < 0.001; $CT_{max}$ – `r cor.test(check_ctmax$trait, check_ctmax$est)$estimate`, 95% CI: `r cor.test(check_ctmax$trait, check_ctmax$est)$conf.int[1]` to `r cor.test(check_ctmax$trait, check_ctmax$est)$conf.int[2]`, p < 0.001; CFLs – `r cor.test(check_cfls$trait, check_cfls$est)$estimate`, 95% CI: `r cor.test(check_cfls$trait, check_cfls$est)$conf.int[1]` to `r cor.test(check_cfls$trait, check_cfls$est)$conf.int[2]`, p < 0.001), suggesting that even this basic model could be used to provide sensible quantitative measures for *Drosophila* species without heat tolerance estimates. While modelling approaches such as these are not a replacement for actual empirical data, they could be used to provide informed judgements on thermal tolerance in the absence of such information, but should be tested (using cross-validation) and refined when necessary.

##### Static lethal limits are evolutionarily correlated with sub-lethal limits in *Drosophilids*
Interestingly, there was a reasonably strong phylogenetic correlation between $CT_{max}$ and critical fertility limits in *Drosophilids* (r = `r posterior.mode(phylocor_ctmax_cfls)`, 95% CI: `r HPDinterval(phylocor_ctmax_cfls)[,"lower"]` to `r HPDinterval(phylocor_ctmax_cfls)[,"upper"]`), but not between $CT_{max}$, CFLs and the slope from TDT curves (cor($CT_{max}$, $z^{TDT}$) = `r posterior.mode(phylocor_ctmax_tdt)`, 95% CI: `r HPDinterval(phylocor_ctmax_tdt)[,"lower"]` to `r HPDinterval(phylocor_ctmax_tdt)[,"upper"]`; cor(CFLs, $z^{TDT}$) = `r posterior.mode(phylocor_tdt_cfls)`, 95% CI: `r HPDinterval(phylocor_tdt_cfls)[,"lower"]` to `r HPDinterval(phylocor_tdt_cfls)[,"upper"]`). These results suggest that sub-lethal measures, such as CFLs strongly coincide with a species' upper lethal limit. It would be interesting to understand if this is a general phenomena across more taxonomic groups.

@@

@@

  layout.matrix <- matrix(c(1, 1, 1, 2,3,4), nrow = 3, ncol = 2) # Phylogeny takes up 3 panels (plot = 1) and the others take up a single panel.

@@

  # Plot 2
  plot(check_z$trait ~ check_z$est, ylab = "TDT Slope (z)", xlab = "Estimated TDT Slope", pch = 16, cex = 3, cex.lab = 1.4)
  abline(a = 0, b = 1, col = "red")
  dev.off()
} else {
  fig <- magick::image_read(here::here("output", "fig", "fig1.pdf"))
  plot(fig)
}
```

@@

We analysed the thermal death time curves for 8 new species of *Drosophila* using the same methods as @Jorgensen2019, using linear models with log 10 coma time as the response and assay temperature as the predcitor (@fig-tdt). We then used the model to estimate the slope and intercept ($CT_{max}$) for each species (Table 1).

>>>>

Phylogenetic comparative approaches could be leveraged to help fill these knowledge gaps because closely related species are often more likely to share similar traits owing to their shared evolutionary history (phylogenetic niche conservatism) [@Losos2008]. If heat tolerance is linked to the phylogeny, shared characteristics that make species more or less heat tolerant will not be randomly distributed across the phylogeny but biased by species groups [@McKinney1997]. We can use evolutionary relatedness to predict species thermal sensitivities and such insights can be used to inform policy and conservation related decisions making for species with little to no data.

@@

##### Evolutionary relationships can help understand thermal tolerance in species lacking data
The model predicted known data reasonably well for most traits (correlation between known values and values estimated from the model: $z^{TDT}$ – `r cor.test(check_z$trait, check_z$est)$estimate`, 95% CI: `r cor.test(check_z$trait, check_z$est)$conf.int[1]` to `r cor.test(check_z$trait, check_z$est)$conf.int[2]`, p < 0.001, @fig-treesub; $CT_{max}$ – `r cor.test(check_ctmax$trait, check_ctmax$est)$estimate`, 95% CI: `r cor.test(check_ctmax$trait, check_ctmax$est)$conf.int[1]` to `r cor.test(check_ctmax$trait, check_ctmax$est)$conf.int[2]`, p < 0.001; CFLs – `r cor.test(check_cfls$trait, check_cfls$est)$estimate`, 95% CI: `r cor.test(check_cfls$trait, check_cfls$est)$conf.int[1]` to `r cor.test(check_cfls$trait, check_cfls$est)$conf.int[2]`, p < 0.001), suggesting that even this basic model could be used to provide sensible quantitative measures for *Drosophila* species without heat tolerance estimates. While modelling approaches such as these are not a replacement for actual empirical data, they could be used to provide informed judgements on thermal tolerance in the absence of such information, but should be tested (using cross-validation) and refined when necessary.

##### Static lethal limits are evolutionarily correlated with sub-lethal limits in *Drosophilids*
Interestingly, there was a reasonably strong phylogenetic correlation between $CT_{max}$ and critical fertility limits in *Drosophilids* (r = `r posterior.mode(phylocor_ctmax_cfls)`, 95% CI: `r HPDinterval(phylocor_ctmax_cfls)[,"lower"]` to `r HPDinterval(phylocor_ctmax_cfls)[,"upper"]`, @fig-treesub), but not between $CT_{max}$, CFLs and the slope from TDT curves (cor($CT_{max}$, $z^{TDT}$) = `r posterior.mode(phylocor_ctmax_tdt)`, 95% CI: `r HPDinterval(phylocor_ctmax_tdt)[,"lower"]` to `r HPDinterval(phylocor_ctmax_tdt)[,"upper"]`; cor(CFLs, $z^{TDT}$) = `r posterior.mode(phylocor_tdt_cfls)`, 95% CI: `r HPDinterval(phylocor_tdt_cfls)[,"lower"]` to `r HPDinterval(phylocor_tdt_cfls)[,"upper"]`). These results suggest that sub-lethal measures, such as CFLs strongly coincide with a species' upper lethal limit. It would be interesting to understand if this is a general phenomena across more taxonomic groups.

@@

data_ct_cfts <- data  %>% group_by(species)  %>% summarise(ctmax_m = mean(ctmax),
                                                          cfts_m = mean(cfts),
                                                   tdt_z_m = mean(tdt_z))  %>% data.frame()

@@

  layout.matrix <- matrix(c(1, 1,2,3), nrow = 2, ncol = 2) # Phylogeny takes up 3 panels (plot = 1) and the others take up a single panel.

@@

  # Plot 2
  plot(check_z$trait ~ check_z$est, ylab = "TDT Slope (z)", xlab = "Estimated TDT Slope", pch = 16, cex = 3, cex.lab = 1.4)
  abline(a = 0, b = 1, col = "red")

  # PLot 3
    plot(ctmax_m ~ cfts_m, data = data_ct_cfts, ylab = "Critcial Thermal Maximum (CTmax)", xlab = "Crictial Fertility Limits (CFLs)", pch = 16, cex = 3, cex.lab = 1.4)

  dev.off()
} else {
  fig <- magick::image_read(here::here("output", "fig", "fig1.pdf"))
  plot(fig)
}
```

@@

We analysed the thermal death time curves for 8 new species of *Drosophila* using the same methods as @Jorgensen2019, using linear models with log 10 coma time as the response and assay temperature as the predictor (@fig-tdt). We then used the model to estimate the slope and intercept ($CT_{max}$) for each species (Table 1).
